---
title       : "Non-Parametric Tests for Differences"
subtitle    : "BIOL40102: Research Methods and Bioethics"
author      : Dr Axel Barlow
job         : "email: axel.barlow@ntu.ac.uk"
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # {zenburn, tomorrow, solarized-dark, ...}
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {selfcontained, standalone, draft}
knit        : slidify::knit2slides
logo        : ntu-shield.png
biglogo     : NTU_open-graph.png
assets      : {assets: ../../assets}
license     : by-nc-sa
github:
  user: draxelbarlow
  repo: BIOL40102_statistics3
  branch: "gh-pages"
---

```{r setup, include = FALSE}
library(tidyverse)
library(gganimate)
library(RColorBrewer)
library(knitr)
library(DiagrammeR)
library(palmerpenguins)
```

<!-- adding bold and italic options -->
<style>
em {
  font-style: italic
}
strong {
  font-weight: bold;
}
</style>

## In this lecture...

  >- Selecting the appropriate test
  >- Non-parametric tests
  >- Mann-Whitney U Test

--- .class #id

## Chi-squared tests

- Can be used for counts of categorical variables

### Chi-squared test of frequencies

- Test if the observed frequencies differ from the expected
- Useful in situations where we can predict frequencies under a particular set or rules or processes

### Chi-squared test of independence

- Test if the observed frequencies between two groups are different
- Expected frequencies can be unknown

--- .class #id

## What about quantitative and ordinal variables?

- Continuous and ordinal variables provide a quantification
- We may want to know if different groups differ in size or magnitude

### Continuous variable, normal distribution

- Parametric tests
- E.g. normally distributed data, 2 groups: t-test (week 32)

### Continuous variable, unknown or complex distribution

- Non-parametric test

### Ordinal variable

- Non-parametric test

--- .class #id 

## Non-parametric vs. parametric tests

- Make fewer assumptions on the underlying data
- Can be a "safer" option if some aspects of the data are unknown
- Lower statistical power (higher false negative rate)

### Used when a parametric test is not appropriate

--- .class #id 

## How does it work?

- Not parametric tests tend to be based on ranks
- Example: how fast can students and professors carry out an experiment?


Student

`10 10 20 21 27 45`

Professor

`15 16 32 33 35 37`

--- .class #id 

## How does it work?

- Not parametric tests tend to be based on ranks
- Example: how fast can students and professors carry out an experiment?

Student

`10 10 -- -- 20 21 27 -- -- -- -- 45`

Professor

`-- -- 15 16 -- -- -- 32 33 35 37 --`

--- .class #id 

## How does it work?

- Not parametric tests tend to be based on ranks
- Example: how fast can students and professors carry out an experiment?


Student

` 1  2 -- --  5  6  7 -- -- -- -- 12 = 33`

Professor

`-- --  3  4 -- -- --  8  9 10 11 -- = 45`

  >- Note the original times have been replaced
  >- So the distribution of the data is not important, just their order
  >- Because we convert the data to ranks, the methods are naturally suited to ordinal data

--- .class #id 

## Mann-Whitney U-Test

- Considered a non-parametric equivalent of the t-test
- 2 groups
- Ordinal or quantitative variable
- Observations must be independent

### Hypotheses

- Do values in one group tend to be larger/smaller than the other group?
- If the distributions are similar: are their medians different?

--- &twocol

## Mann-Whitney U-Test

Example: oxygen saturation for 2 groups of patients (N= 100 per group)

```{r, echo = FALSE}

sample1 <- enframe(rbeta(100,100,1))
sample1 <- sample1 %>% mutate(O2 = round((value * 100) - 1)) %>% add_column(group = 1)
#write.csv(sample1, file="O2_group1.txt")

sample2 <- enframe(rbeta(100,100,1))
sample2 <- sample2 %>% mutate(O2 = round((value * 99) - 1)) %>% add_column(group = 2)
#write.csv(sample1, file="O2_group2.txt")

plot1 <- ggplot(sample1) +
  geom_histogram(aes(x=O2), fill="wheat", col="black", position="identity", show.legend=TRUE, bins = 6) +
  theme_bw(base_size = 20) +
  xlab("fragment length (bp)")

#plot1

plot2 <- ggplot(sample2) +
  geom_histogram(aes(x=O2), fill="lightgreen", col="black", position="identity", show.legend=TRUE, bins = 6) +
  theme_bw(base_size = 20) +
  xlab("fragment length (bp)")

#plot2

total <- rbind(sample1, sample2)

utest <- wilcox.test(O2 ~ group, data=total)
```
  
*** =left

**Group1**

```{r, echo=FALSE}
print(sample1[3])
```

*** =right

**Group2**

```{r, echo=FALSE}
print(sample2[3])
```

--- &twocol bg:white

## Mann-Whitney U-Test

Example: oxygen saturation for 2 groups of patients (N= 100 per group)

*** =left

**Group1**

```{r, out.width = "100%", fig.width = 6, fig.height = 5, dpi = 600, fig.align = "left", echo = FALSE}
plot1
```

*** =right

**Group2**

```{r, out.width = "100%", fig.width = 6, fig.height = 5, dpi = 600, fig.align = "left", echo = FALSE}
plot2
```

--- .class #id

## Mann-Whitney U-Test

Example: oxygen saturation for 2 groups of patients (N= 100 per group)

- Median Group 1 `r format(round(median(sample1$O2), 0))`% O2
- Median Group 2 `r format(round(median(sample2$O2), 0))`% O2
- Mann-Whitney U-test results: W = `r format(round(utest$statistic, 2))`, p-value = `r utest$p.value`

--- &thankyou

## Next time

**What do you need to know?**